---
title: "Reservoir Mass Balance Simulation with Package `outfloweR`"
author: "Thomas Petzoldt"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: bib.bib
vignette: >
  %\VignetteIndexEntry{Reservoir Mass Balance Simulation with Package `outfloweR`}
  %\VignetteEngine{knitr::rmarkdown}
---
<!-- avoid border around images -->
<style>img {border: 0;}</style>


```{r opts, echo = FALSE, message = FALSE}
library("dplyr")
library("ggpubr")
library("outfloweR")
knitr::opts_chunk$set(fig.width=6, fig.height=4)
``` 

# Introduction

The package aims to demonstrate water management alternatives by means of a simple
mass balance model. Currently, the following state variables and inputs are considered:

## State variables

* Volume of the mixed layer ($V_E$), epilimnion in summer, total volume otherwise)
* Volume of the hypolimnion ($V_H$)

### State equations

\begin{align*}
\frac{dV_E}{dt} &= Q_{in, E} - Q_{out, E} - Q_{out, wb, E} \\
\frac{dV_H}{dt} &= Q_{in, H} - Q_{out, H} - Q_{out, wb, H}
\end{align*}

with:

$$
dV_E = dV_E + dV_H \qquad | \qquad V_H = 0
$$

The start time of summer stratification $t_{strat}$ can be set by the user. At this time point, the total volume $V$ is subdivided into an epilimnetic volume  $V_E = V \cdot f_{epi}$ and a hypolimnetic part $V_H = V \cdot (1- f_{epi})$. The end of summer stratification occurs when the hypolimnion is exhausted ($V_H = 0$) or time exceeds
fixed a threshold ($t_{autumn$), mimicking thermal convection. This simple approach 
was intentionally chosen, even if a full hydrophysical 1D model is also available.


## Inputs

* Considered are water inflow (`inflow`), outflow (`outflow`) and wild bed outlet (`outflow_wb`).
* The water inflow can entrain to the epilimnion or the hypolimnion. If a hypolimnion does not exist, hypolimnetic inflows are added to the epilimnion.


# Scenarios

Water inflow and abstraction can be controlled by pre-defined scenarios, currently the following:

* standard: epilimnetic inflow, hypolimnetic outflows
* cold_inflow: hypolimnetic inflow and outflows
* wb_top: epilimnetic inflow, hypolimnetic drinking water abstraction, wild bed outflow taken from epilimnion

Further important settings are `vol`, the initial volume of the reservoir and `t_strat` the time of stratification onset. The parameter `t_autumn` specifies the latest end date
of stratificatiion by thermal convection, but stratification can also occur earlier
if the hypolimnion volume is exhausted.

# Examples

## Series 1: realistic time series with balanced inflow and outflow

The first example shows the standard example with the `discharge_normal` example data set
provided by the package.

```{r, fig.height=6}
library("outfloweR")
data("discharge_normal")
plot_flows(discharge_normal)
```

As a first attempt, we assume that the inflow is rather cold, so that it entrains to the hypolimnion. All outflows are also abstracted from the bottom.

```{r, fig.width=7.5}
out <- scenario("cold_inflow", vol=60, t_strat=90, t_autumn=300,
                discharge_normal$inflow, discharge_normal$outflow,
                discharge_normal$outflow_wb)

plot_volumes(out)
```

In the given data example, inflow and outflow data are well balanced, so that the hypolimnion remains at an approximately constant level. This is of course an 
approximation, ignoring further hydrophysical processes.

Moreover, a cold hypolimnetic inflow is rarely observed in nature. In most cases, the 
inflowing water entrains to the epi- or metalimnion. A metalimnion is not
considered here for simplicity, so we assume an entrainment to the mixed layer.

```{r, , fig.width=7.5}
out <- scenario("standard", vol=60, t_strat=90, t_autumn=300,
                discharge_normal$inflow, discharge_normal$outflow,
                discharge_normal$outflow_wb)

plot_volumes(out)
```

We see that the hypolimnion is exhausted over the year.


The next example shows what happens if the wild bed outflow is abstracted from
the epilimnion:

```{r, , fig.width=7.5}
out <- scenario("wb_top", vol=60, t_strat=90, t_autumn=300,
                discharge_normal$inflow, discharge_normal$outflow,
                discharge_normal$outflow_wb)

plot_volumes(out)
```

We see that hypolimnetic volume can be saved, i.e. we have more volume available 
volume for drinking water abstraction. This will also have positive effects on water quality.

## Example 2: Simplified time series with a summer flood

Here we create an artifical inflow and outflow time series, where we assume
constant inflows and outflows, plus a summer flood event generated from a 
log-lormal function. This leads to an increased total volume, that is then
subsequently removed from the dam.

The data creation is somewhat technical and is given here for full reproducibility.
It may be integrated as example data set in the package at a later time.

```{r}
library("dplyr")
const_discharge <- data.frame(
  time = 1:365,
  inflow = 0.2,
  outflow=0.1,
  outflow_wb=0.12
)
time <- 1:365; wave_start <- 150; yscale <- 10
wave <- yscale * dlnorm(time - wave_start, meanlog=2, sdlog=1)

summer_flood <- const_discharge |>
  mutate(inflow = inflow + wave,
         outflow_wb = ifelse(wave_start + 20 < time & time <= wave_start + 40, outflow + sum(wave)/20, outflow))

plot_flows(summer_flood)
```

Now let's compare the mass balance model with scnearios `standard` (hypolimnetic wild bed) and `wb_top` (wild bed abstraction from epilimnion):

```{r}
library("ggpubr")
out1 <- with(summer_flood,
            scenario("standard", vol=60, t_strat=90, t_autumn=300,
              inflow, outflow, outflow_wb))

out2 <-  with(summer_flood,
              scenario("wb_top", vol=60, t_strat=90, t_autumn=300,
                       inflow, outflow, outflow_wb))
fig1 <- plot_volumes(out1)
fig2 <- plot_volumes(out2)                        
ggpubr::ggarrange(fig1, fig2, ncol=1)
```

In result, we can clearly see the loss of hypolimnetic volume in the standard 
scenario and the shorter stratification period (top figure) and the extra hypolimnion volume gained in the second scenario, if wild bed outled is taken from the epilimnetic layer.

